Air Quality Data - PM2.5 Components

### 2021-01-22 code/data-analysis/factor-analysis.Rmd
PMC and CMS Merge Config file generator factor analysis output data cleaning forest plot

### This program:
  (1) Run factor anlaysis

```{r load packages}
library(tidyverse)
library(here)
library(psych) # for factor analysis

```

```{r Load datasets}
load(here('data','processed', 'pmc-selected.RDa'))
```

```{r - preliminary factor analysis}
# All eligible monitors
fit <- factanal(dt.pmc.sel %>% select(AS:OC), 8, rotation="varimax")
factor.loadings <- rbind(fit$loadings)
save(factor.loadings, file = here('output','table','factor-loadings.RDa'))

factor.scores <- factor.scores(dt.pmc.sel %>% select(AS:OC), fit)
dt.pmc.fac <- cbind(dt.pmc.sel, factor.scores$scores)
save(dt.pmc.fac, file = here('data', 'processed', 'pmc-fac.RDa'))

# Removing EC from factor analysis
fit <- factanal(dt.pmc.sel %>% select(AS:ZN, OC), 8, rotation="varimax")
factor.loadings.ex.ec <- rbind(fit$loadings)
save(factor.loadings.ex.ec, file = here('output','table','factor-loadings-ex-ec.RDa'))

factor.scores <- factor.scores(dt.pmc.sel %>% select(AS:ZN, OC), fit)
dt.pmc.fac.ex.ec <- cbind(dt.pmc.sel, factor.scores$scores)
save(dt.pmc.fac.ex.ec, file = here('data', 'processed', 'pmc-fac-ex-ec.RDa'))
```

>>>>

### This program:
  (1) Run factor analysis

```{r load packages}
library(tidyverse)
library(here)
library(psych) # for factor analysis
```

```{r Load datasets}
load(here('data','processed', 'pmc-daily.RDa'))
```

```{r Call functions}
source(here('code','data-processors','air-data-smooth-mvavg-function.R'))
```

```{r - preliminary factor analysis}
# All eligible monitors
fit <- factanal(dt.pmc.daily %>% select(AS:OC), 8, rotation="varimax")
factor.loadings <- rbind(fit$loadings)
save(factor.loadings, file = here('output','table','factor-loadings.RDa'))
#Traffic Soil Metals Biomass Salt Oil Coal Steel

factor.scores <- factor.scores(dt.pmc.daily %>% select(AS:OC), fit)
dt.fac.daily <- cbind(dt.pmc.daily %>% select(Site.ID, Date), factor.scores$scores) 
save(dt.fac.daily, file = here('data', 'processed', 'fac-daily.RDa'))
```

```{r Removing EC from factor analysis}
fit <- factanal(dt.pmc.daily %>% select(AS:ZN, OC), 8, rotation="varimax")
factor.loadings.ex.ec <- rbind(fit$loadings)
save(factor.loadings.ex.ec, file = here('output','table','factor-loadings-ex-ec.RDa'))
# Soil Metals Traffic Biomass Coal Salt Oil Steel

factor.scores <- factor.scores(dt.pmc.daily %>% select(AS:ZN, OC), fit)
dt.fac.daily.ex.ec <- cbind(dt.pmc.daily %>% select(Site.ID, Date), factor.scores$scores)
save(dt.fac.daily.ex.ec, file = here('data', 'processed', 'fac-daily-ex-ec.RDa'))
```

```{r Smoothing Data & Caluculating 1-yr mvavg for each month}
dt.air <- dt.fac.daily %>% 
  pivot_longer(cols = Factor1:Factor8,
               names_to = "Parameter.Symbol",
               values_to = "Value") %>% 
  arrange(Site.ID, Parameter.Symbol, Date)

site.para <- dt.air %>% 
  distinct(Site.ID, Parameter.Symbol)

#using smoothed.yr function
smooth.out <- mapply(Smoothed.Year, site.para$Site.ID, site.para$Parameter.Symbol)
para.monthly <- data.frame(Site.ID=unlist(smooth.out[1,]),
                           Parameter.Symbol=unlist(smooth.out[2,]),
                           Date=unlist(smooth.out[3,]), 
                           para.1yr=unlist(smooth.out[4,]))

dt.fac.1yr <- para.monthly %>% 
  mutate(Year = floor(Date -0.01),
         Month = round((Date - Year)*12)) %>% 
  select(-Date) %>% 
  pivot_wider(names_sort = T,
              names_from = c(Parameter.Symbol),
              values_from = c(para.1yr)) %>% 
  arrange(Site.ID, Year, Month)

save(dt.fac.1yr, file = here('data', 'processed', 'fac-1yr.RDa'))
write.csv(dt.fac.1yr, file = here('data', 'processed', 'fac-1yr.csv'), row.names = F)
```

```{r Smoothing Data & Caluculating 1-yr mvavg for each month}
dt.air <- dt.fac.daily.ex.ec %>% 
  pivot_longer(cols = Factor1:Factor8,
               names_to = "Parameter.Symbol",
               values_to = "Value") %>% 
  arrange(Site.ID, Parameter.Symbol, Date)

site.para <- dt.air %>% 
  distinct(Site.ID, Parameter.Symbol)

#using smoothed.yr function
smooth.out <- mapply(Smoothed.Year, site.para$Site.ID, site.para$Parameter.Symbol)
para.monthly <- data.frame(Site.ID=unlist(smooth.out[1,]),
                           Parameter.Symbol=unlist(smooth.out[2,]),
                           Date=unlist(smooth.out[3,]), 
                           para.1yr=unlist(smooth.out[4,]))

dt.fac.ex.ec.1yr <- para.monthly %>% 
  mutate(Year = floor(Date -0.01),
         Month = round((Date - Year)*12)) %>% 
  select(-Date) %>% 
  pivot_wider(names_sort = T,
              names_from = c(Parameter.Symbol),
              values_from = c(para.1yr)) %>% 
  arrange(Site.ID, Year, Month)

save(dt.fac.ex.ec.1yr, file = here('data', 'processed', 'fac-ex-ec-1yr.RDa'))
write.csv(dt.fac.ex.ec.1yr, file = here('data', 'processed', 'fac-ex-ec-1yr.csv'), row.names = F)
```

### 2021-01-22 code/data-processors/air-data-processor.Rmd
PMC and CMS Merge Config file generator factor analysis output data cleaning forest plot

```{r load packages}
library(tidyverse)
library(here)
library(lubridate)
library(naniar) # for replacing -999 with NA
```

@@

save(dt.pmc.site.all.zip.bz12, file = here('data', 'processed', 'pmc-site-all-zip-bz12.RDa'))
write.csv(dt.pmc.site.all.zip.bz12, file = here('data', 'processed', 'pmc-site-all-zip-bz12.csv'))

@@

save(dt.pmc.1yr, file = here('data', 'processed', 'pmc-1yr.RDa'))
write.csv(dt.pmc.1yr, file = here('data', 'processed', 'pmc-1yr.csv'))

dt.site.pmc <- dt.pmc.1yr %>% distinct(Site.ID) %>%  #574 -> 558 > 471 > 318 > 318
  merge(dt.aqs.sites)
save(dt.site.pmc, file = here('data', 'processed', 'site-pmc.RDa'))

dt.site.pmc %>% 
  group_by(Location.Setting) %>% 
  summarise(n = n())
#Location.Setting  n
#                  2			
#RURAL	           177			
#SUBURBAN	         67			
#URBAN	           72

dt.pmc.1yr %>% 
  distinct(Site.ID, Year) %>% 
  group_by(Site.ID) %>% 
  summarise(n = n()) %>% filter(n > 3) %>% 
  nrow() #318: all  #311: with 4+ years

@@

summary(check$n)
```

@@

save(dt.pmc.site.zip.bz6, file = here('data', 'processed', 'pmc-site-zip-bz6.RDa'))
write.csv(dt.pmc.site.zip.bz6, file = here('data', 'processed', 'pmc-site-zip-bz6.csv'))

# Count number of zipcodes and sites
dt.pmc.site.zip.bz6 %>% distinct(Zip.Code) %>% nrow() #4253
dt.pmc.site.zip.bz6 %>% distinct(Site.ID) %>% nrow() #318 > 270

@@

save(dt.pmc.site.zip.bz12, file = here('data', 'processed', 'pmc-site-zip-bz12.RDa'))
write.csv(dt.pmc.site.zip.bz12, file = here('data', 'processed', 'pmc-site-zip-bz12.csv'))

# Count number of zipcodes and sites
dt.pmc.site.zip.bz12 %>% distinct(Zip.Code) %>% nrow() #8313
dt.pmc.site.zip.bz12 %>% distinct(Site.ID) %>% nrow() #318 > 316

@@

save(dt.pmc.site.zip.bz24, file = here('data', 'processed', 'pmc-site-zip-bz24.RDa'))
write.csv(dt.pmc.site.zip.bz24, file = here('data', 'processed', 'pmc-site-zip-bz24.csv'))

# Count number of zipcodes and sites
dt.pmc.site.zip.bz24 %>% distinct(Zip.Code) %>% nrow() #15601
dt.pmc.site.zip.bz24 %>% distinct(Site.ID) %>% nrow() #318 > 317

# pmc-site-zip-bz12.csv was uploaded on NEU Cluster to be merged with CMS data
# It has 316 sites with 8313 zipcodes around them
```

```{r Post mergeing with CMS Data}
dt.site.ndi.bz12 <- read.csv(here('data', 'raw', 'site_pmc_ndi_bz12.csv')) %>% 
  rename(Site.ID = Site_ID)

# 325 out of 328 sites have population living in their 12-mile bufferzone
dt.site.ndi.bz12 %>% 
  group_by(Site.ID) %>% 
  summarise(n = n()) %>% filter(n > 0) %>% 
  nrow() # 325 > 317 > 316 > 308: with 4+ years

# TOT/TOR Method in different location.set
dt.site.ndi.bz12 %>% 
  distinct(Site.ID) %>% 
  left_join(dt.site.carbon.method) %>% 
  left_join(dt.aqs.sites) %>% 
  group_by(Location.Setting, Method) %>% 
  summarise(n = n())

@@

dt.site.ndi.bz12 %>% 
  distinct(Site.ID) %>% 
  left_join(dt.aqs.sites) %>% 
  group_by(Location.Setting) %>% 
  summarise(n = n())
# urban 72
# suburban 69
# rural 182
```

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load packages}
library(tidyverse)
library(here)
#library(lubridate)
#library(naniar) # for replacing -999 with NA
```

@@

save(dt.pmc.site.all.zip.bz12, file = here('data', 'processed', 'pmc-site-all-zip-bz12.RDa'))
write.csv(dt.pmc.site.all.zip.bz12, file = here('data', 'processed', 'pmc-site-all-zip-bz12.csv'), row.names = F)

@@

save(dt.pmc.1yr, file = here('data', 'processed', 'pmc-1yr.RDa'))
write.csv(dt.pmc.1yr, file = here('data', 'processed', 'pmc-1yr.csv'), row.names = F)

dt.site.pmc <- dt.pmc.1yr %>% distinct(Site.ID) %>%  #574 -> 558 > 471 > 322
  merge(dt.aqs.sites)
save(dt.site.pmc, file = here('data', 'processed', 'site-pmc.RDa'))

dt.site.pmc %>% 
  group_by(Location.Setting) %>% 
  summarise(n = n())
#Location.Setting  n
#                  2			
#RURAL	           178			
#SUBURBAN	         69		
#URBAN	           73

dt.pmc.1yr %>% 
  distinct(Site.ID, Year) %>% 
  group_by(Site.ID) %>% 
  summarise(n = n()) %>% filter(n > 3) %>% 
  nrow() #322: all  #315: with 4+ years

@@

summary(check$n) # 23 to 97
```

@@

save(dt.pmc.site.zip.bz6, file = here('data', 'processed', 'pmc-site-zip-bz6.RDa'))
write.csv(dt.pmc.site.zip.bz6, file = here('data', 'processed', 'pmc-site-zip-bz6.csv'),
          row.names = F)

# Count number of zipcodes and sites
dt.pmc.site.zip.bz6 %>% distinct(Zip.Code) %>% nrow() #4348
dt.pmc.site.zip.bz6 %>% distinct(Site.ID) %>% nrow() #322 > 274

@@

save(dt.pmc.site.zip.bz12, file = here('data', 'processed', 'pmc-site-zip-bz12.RDa'))
write.csv(dt.pmc.site.zip.bz12, file = here('data', 'processed', 'pmc-site-zip-bz12.csv'),
          row.names = F)

# Count number of zipcodes and sites
dt.pmc.site.zip.bz12 %>% distinct(Zip.Code) %>% nrow() #8446
dt.pmc.site.zip.bz12 %>% distinct(Site.ID) %>% nrow() #322 > 320

@@

save(dt.pmc.site.zip.bz24, file = here('data', 'processed', 'pmc-site-zip-bz24.RDa'))
write.csv(dt.pmc.site.zip.bz24, file = here('data', 'processed', 'pmc-site-zip-bz24.csv'),
          row.names = F)

# Count number of zipcodes and sites
dt.pmc.site.zip.bz24 %>% distinct(Zip.Code) %>% nrow() #15778
dt.pmc.site.zip.bz24 %>% distinct(Site.ID) %>% nrow() #322 > 321

# pmc-site-zip-bz12.csv was uploaded on NEU Cluster to be merged with CMS data
# It has 320 sites with 8446 zipcodes around them
```

```{r Post mergeing with CMS Data}
dt.pmc.site.ndi <- read.csv(here('data', 'processed', 'site_pmc_ndi.csv')) %>% 
  rename(Site.ID = Site_ID) %>% 
  left_join(dt.aqs.sites) %>% 
  left_join(dt.site.carbon.method)
# 317 out of 320 sites have population living in their 12-mile bufferzone

save(dt.pmc.site.ndi, file = here('data', 'processed', 'pmc-site-ndi.RDa'))
write.csv(dt.pmc.site.ndi, file = here('data', 'processed', 'pmc-site-ndi.csv'),
          row.names = F)

# TOT/TOR Method in different location.set
dt.pmc.site.ndi %>% 
  group_by(Location.Setting, Method) %>% 
  summarise(n = n())

@@

dt.pmc.site.ndi %>% 
  group_by(Location.Setting) %>% 
  summarise(n = n())
# urban 72
# suburban 69
# rural 174
# ...   2
```

```{r}
dt.pmc.daily <- dt.pmc.site.ndi %>% 
  select(Site.ID) %>% 
  merge(dt.pmc.sel)

save(dt.pmc.daily, file = here('data', 'processed', 'pmc-daily.RDa'))
```

```{r}
test <- dt.pmc.daily %>% 
  group_by(Site.ID) %>% 
  nest() %>% 
  mutate(model = map(data, ~lm(PM ~ AS, data = .)),
         resid = map(model, residuals)) %>% 
  unnest(data, resid)
#208037
test2 <- unlist(unlist(test$model))
test2 <- residuals(test$model[[3]])

d <- dt.pmc.daily %>% 
  filter(Site.ID == "01-073-0023")

model <- lm(data = d, PM ~AS)

residuals(model)

d <- data.frame(state=rep(c('NY', 'CA'), c(10, 10)),
                                 year=rep(1:10, 2),
                                 response=c(rnorm(10), rnorm(10)))

d %>% 
  group_by(state) %>% 
  nest() %>% 
  mutate(model = map(data, ~lm(response ~ year, data = .)))
```
